Staged Abstract Interpreters

### 2019-06-30 paper/10_intro.tex
intro

In recent years, we observe a rich progress on how we can construct abstract
interpreters from systematic principles, instead of ad-hoc engineerings. A
notable one is the Abstracting Abstract Machine (AAM) methodology
\cite{DBLP:journals/jfp/HornM12, DBLP:conf/icfp/HornM10}, which uncovers an
approach to derive sound abstract interpreters from their concrete counterparts,
where the soundness can be easily established by examining the transformation of
semantic artifacts. For example, a CEK machine
\cite{DBLP:conf/popl/FelleisenF87} for concrete execution can be easily
refactored to an effective 0-CFA control-flow analysis
\cite{Shivers:1988:CFA:53990.54007, Midtgaard:2012:CAF:2187671.2187672} by first
tweaking the environment dereference as a nondeterministic choice, then
allocating continuations in the environment, and constraining the addresses
space to be finite. This systematic abstraction approach can be tailored to
different analyses \todo{cite} and and has been applied to multiple variants of
small-step abstract machines \cite{DBLP:journals/jfp/HornM12,
DBLP:conf/icfp/HornM10, Sergey:2013:MAI:2491956.2491979} and big-step
definitional interpreters \cite{Wei:2018:RAA:3243631.3236800,
DBLP:journals/pacmpl/DaraisLNH17, Keidel:2018:CSP:3243631.3236767}.

More pragmatically, as an implementation strategy, several purely functional
programming approaches to build abstract interpreters were emerged, for
examples, using monads or arrows. The pure approach provides much benefits: 1)
the abstract interpretation artifacts are compositional and can be constructed
modularly, 2) one can easily prove the soundness of analysis, and 3) one can
easily reason the correctness of implementation by leveraging equational
reasoning. For example, monadic abstract interpreters \todo{two level semantics,
arrow, Galois transformers}

However, besides the intrinsic complexity of static analysis, there are
additional abstraction penalties with these high-level programming approaches.
First, similar to concrete interpreters, the abstract interpreter analyzes the
program by traversing the abstract syntax tree, which poses the interpretive
overhead, e.g., pattern matching on the AST and recursive calls on the sub
expressions. Those kind of overhead can be negligible if the abstract
interpreter only runs on the program once, but also can be accumulated
significantly if it runs repeatedly, for example, on libraries. Second, the
abstract interpreter written in pure languages usually extensively uses effect
systems to model the behaviors of abstract interpretatation, such as
nondeterminism. For example, \citet{DBLP:journals/pacmpl/DaraisLNH17} and
\citet{Sergey:2013:MAI:2491956.2491979}'s monadic abstract interpreters use
monad transformers; the very recent \citet{Githubsemantic}'s Semantic framework
uses extensible effects. The effect systems may introduce additional performance
overhead.

In this paper, we propose an abstraction-without-regret approach to eliminate
those performance penalties for abstract interpreters written in purely
functional programming style. In short, our approach borrows ideas from program
specialization and high-performance DSLs: 1) Using multi-stage programming at
type level (the Lightweight Modular Staging framework
\cite{DBLP:conf/gpce/RompfO10}, in this paper), we can
specialize the abstract interpreter with respect to the input program and then
generate efficient low-level code that does the actual analysis. 2) We use the
tagless-final approach to embed the target language interpreter in a high-level
host language (Scala, in this paper), allowing the user to implement different
semantics modularly, including the abstract semantics, without changing the main
generic interpreter. This design choice allows user to derive staged abstract
interpreters without intrusive changes to the unstaged one. Therefore, our
approach is no regret in the sense of both performance and least engineering
efforts to achieve such performance. We elaborate these two main ideas in detail
as follow.

@@

@@

>>>>

In recent years, we observe a rich progress on how we can construct
abstract interpreters from systematic principles, instead of ad-hoc
engineerings. A notable one is the Abstracting Abstract Machine (AAM)
methodology \cite{DBLP:journals/jfp/HornM12, DBLP:conf/icfp/HornM10},
which uncovers an approach to derive sound abstract interpreters from
their concrete counterparts, where the soundness can be easily
established by examining the transformation of semantic artifacts. For
example, a CEK machine \cite{DBLP:conf/popl/FelleisenF87} for concrete
execution can be easily refactored to an effective 0-CFA control-flow
analysis \cite{Shivers:1988:CFA:53990.54007, Midtgaard:2012:CAF:2187671.2187672}
by first tweaking the environment dereference as a nondeterministic choice,
then allocating continuations in the environment, and constraining the
addresses space to be finite. This systematic abstraction approach can
be tailored to different analyses \todo{cite}
\cite{DBLP:conf/icfp/Gilray0M16, DBLP:conf/popl/GilrayL0MH16} and and
has been applied to multiple variants of small-step abstract machines
\cite{DBLP:journals/jfp/HornM12, DBLP:conf/icfp/HornM10,
  Sergey:2013:MAI:2491956.2491979} and big-step definitional
interpreters \cite{Wei:2018:RAA:3243631.3236800,
  DBLP:journals/pacmpl/DaraisLNH17, Keidel:2018:CSP:3243631.3236767}.

More pragmatically, as an implementation strategy, several purely
functional programming approaches to build abstract interpreters were
emerged, for examples, using monads or arrows. The pure approach
provides much benefits: 1) the abstract interpretation artifacts are
compositional and can be constructed modularly, e.g., by using
monad transformers. 2) The soundness of analysis can be proved more easily, by
mechnization \cite{Darais:2016:CGC:2951913.2951934} or paper-based
proof \cite{Keidel:2018:CSP:3243631.3236767}.  3) The correctness of
implementations can be reasoned by the programmers more condidently,
e.g., by leveraging equational reasoning.

However, besides the intrinsic complexity of static analysis, there are
additional abstraction penalties with these high-level programming approaches.
First, similar to concrete interpreters, the abstract interpreter analyzes the
program by traversing the abstract syntax tree, which poses the interpretive
overhead, e.g., pattern matching on the AST and recursive calls on the sub
expressions. Those kind of overhead can be negligible if the abstract
interpreter only runs on the program once, but also can be accumulated
significantly if it runs repeatedly, for example, on libraries. Second, the
abstract interpreter written in pure languages usually extensively uses effect
systems to model the behaviors of abstract interpretatation, such as
nondeterminism. For example, \citet{DBLP:journals/pacmpl/DaraisLNH17} and
\citet{Sergey:2013:MAI:2491956.2491979}'s monadic abstract interpreters use
monad transformers; the very recent \citet{Githubsemantic}'s \textsc{Semantic} framework
uses extensible effects. With certain merits and elegance of the implementation,
the effect systems may also introduce additional performance overhead.

In this paper, we propose an abstraction-without-regret approach to
eliminate those performance penalties for abstract interpreters
written in purely functional programming style. In short, our approach
borrows ideas from program specialization and high-performance DSLs,
and applies them to a particular kind of meta-programs, abstract
interpreters.
1) Using multi-stage programming (the Lightweight Modular Staging
framework \cite{DBLP:conf/gpce/RompfO10}, in this paper), we can
specialize the abstract interpreter with respect to an input program
and then generate efficient low-level code that does the actual
analysis. The specialization can also eliminate the effect layers in the generated code.
2) We use a tagless-final approach to embed the target
language interpreter in a high-level host language (Scala), allowing
the user to implement different semantics modularly, including the
abstract semantics and staging, without changing the main generic
interpreter. Together with the multi-stage programming and type-level
annotations, this design allows user to derive staged abstract
interpreters without intrusive changes to the unstaged one. Therefore,
our approach is no regret in the sense of both performance and least
engineering efforts to achieve such performance. We elaborate these
two main ideas in detail as follow.

@@

The idea of

@@

\paragraph{Applications}

\paragraph{Contribution}

\paragraph{Oragnization}
